---
title: "project1_code"
author: "Tianyu Wu, Yue Li"
date: "2022-09-18"
output: html_document
---

### This is the R code demo for project 1 of STA 521, Fall 2022 ###


```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(gridExtra)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
log_data <- read_csv("~/sta 521 proj1/sonoma-data-log.csv")
net_data <- read_csv("~/sta 521 proj1/sonoma-data-net.csv")
loc_data <- read.table("~/sta 521 proj1/mote-location-data.txt", header = T, sep = "")
```

```{r}
date0 = read.delim('~/sta 521 proj1/sonoma-dates', header = F)
# epochNums
epochNums = substring(date0[1,1], 14, nchar(date0[1,1])-3)
epochNums = strsplit(epochNums, " ")
epochNums = as.numeric(unlist(epochNums))
# epochDates
epochDates = substring(date0[2,1], 16, nchar(date0[2,1])-4)
epochDates = strsplit(epochDates, "\' \'")
epochDates = as.POSIXlt(unlist(epochDates), format = "%a %b  %d %H:%M:%S %Y")
# epochDays
epochDays = substring(date0[3,1], 14, nchar(date0[3,1])-3)
epochDays = strsplit(epochDays, " ")
epochDays = as.numeric(unlist(epochDays))
# create the new data frame
new_date0 = data.frame(epochNums, epochDates, epochDays)
colnames(new_date0)[1] = "epoch"
colnames(new_date0)[2] = "result_time"
colnames(new_date0)[3] = "result_days"
```

# Part 2 Data Cleaning

```{r}
nrow(log_data)
log_data <- merge(log_data, new_date0, by = "epoch") |>
  mutate(result_time = result_time.y) |>
  select(-result_days, -ends_with(".x"), -ends_with(".y"))
nrow(log_data)
```

```{r}
nrow(net_data)
net_data <- merge(net_data, new_date0, by = "epoch") |>
  mutate(result_time = result_time.y) |>
  select(-result_days, -ends_with(".x"), -ends_with(".y"))
nrow(net_data)
```

## (a)

First use summary to extract useful insights from summary statistics.

```{r}
variables = c("humidity", "humid_temp", "hamatop", "hamabot", "voltage")
log_data |>
  select(all_of(variables)) |>
  summary()
```

```{r}
net_data |>
  select(all_of(variables)) |>
  summary()
```

From the summary statistics we know that 
1. Variable voltage for two datasets are not in the same domain.
2. There are NAs for all of the datasets.

We know from the paper that the incident and reflected PAR measurements were collected by two Hamamatsu S1087 photodiodes interfaced to the 10-bit
ADC on Mica2Dot. From user manual (http://www-db.ics.uci.edu/pages/research/quasar/MPR-MIB%20Series%20User%20Manual%207430-0021-06_A.pdf Page 25) we found the formula of conversion to be $V_{batt} = V_{ref} \times ADC\_FS / ADC\_Count$. The voltage data from net dataset is converted to the right range after the conversion: 

```{r}
# 
ADC_FS = 1024
V_ref = 0.6 

net_data |>
  select(voltage) |>
  mutate(voltage_battery = V_ref * ADC_FS / voltage) |>
  summary()
```

To make sure the conversion is reasonable, plot the histogram of both voltage columns after removing NAs and outliers from eye-ball checking (Outlier detection and removal is covered in detail in part 2(d)).

```{r}
# Variable voltage
plot1 <- log_data |>
           ggplot(aes(x = voltage)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data |>
           ggplot(aes(x = voltage)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

```{r}
# Variable voltage
plot1 <- log_data |>
           filter(voltage >= 2) |>
           ggplot(aes(x = voltage)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data |>
           filter(voltage <= 400) |>
           mutate(voltage_battery = V_ref * ADC_FS / voltage) |>
           ggplot(aes(x = voltage_battery)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

We can safely replace original voltage with the transformed values.

```{r}
net_data <- net_data |>
              mutate(voltage = V_ref * ADC_FS / voltage)
```

## (b)

It turns out that when a rows contains any NA, all four variables of interests are NA:

```{r}
colSums(is.na(log_data))
sum(apply(log_data, 1, anyNA))
colSums(is.na(net_data))
sum(apply(net_data, 1, anyNA))
```

Identify corresponding date and time period:

```{r}
log_data_na <- log_data[rowSums(is.na(log_data)) > 0,]
log_data_na |> 
  select(result_time) |>
  distinct()
```

However, note that the log data has the same result_time. Therefore, we should consider using epoch.

```{r}
log_data |>
  select(result_time) |>
  distinct()
```

```{r}
log_data_na |> 
  select(epoch) |>
  distinct() |>
  arrange(epoch)
```

Most of the time there is only one NA for each epoch.

```{r}
log_data_na |> 
  select(epoch) |>
  ggplot(aes(x = epoch)) +
  geom_bar()
```

The distribution of NAs are listed below:

```{r}
log_data_na |> 
  group_by(nodeid) |>
  dplyr::summarise(
    count = n(),
    start = min(epoch),
    end = max(epoch),
    span = max(epoch) - min(epoch) + 1)
```

Number of rows after distinct() is the same as number of NA. In other words, all result_time are different.

```{r}
net_data_na <- net_data[rowSums(is.na(net_data)) > 0,]
net_data_na |> 
  select(result_time) |>
  distinct()
```

Examine more on row 9 and row 10 as they are the same when displayed above.

```{r}
net_data_na_time <- net_data_na |> 
                      select(result_time) |>
                      distinct()
na_result_time <- net_data_na_time$result_time
na_result_time[9] - na_result_time[10]
```

It turns out that the difference is in seconds. Therefore, we can round the date time to minutes level. As can be seen below, 3756/4262 rows remain. All of them indeed have different date and time.

```{r}
net_data_na |> 
  select(result_time) |>
  mutate(result_time = as.POSIXct(round(result_time, "mins"))) |>
  distinct()
```

514/4262 rows remain if round to hours

```{r}
net_data_na |> 
  select(result_time) |>
  mutate(result_time = as.POSIXct(round(result_time, "hours"))) |>
  group_by(result_time) |>
  dplyr::summarise(count = n())
  # dplyr::count(result_time) equivalent
```

```{r}
net_data_na |> 
  select(result_time) |>
  mutate(result_time = as.POSIXct(round(result_time, "hours"))) |>
  ggplot(aes(x = result_time)) +
  geom_bar()
```

The time spans 22 days from May 08 to May 29 (everyday in this span).

```{r}
net_data_na |> 
  select(result_time) |>
  mutate(result_time = as.POSIXct(round(result_time, "days"))) |>
  dplyr::count(result_time)
```

```{r}
net_data_na |> 
  select(result_time) |>
  mutate(result_time = as.POSIXct(round(result_time, "days"))) |>
  ggplot(aes(x = result_time)) +
  geom_bar()
```

Remove missing values and do sanity check.

```{r}
nrow_log_data <- nrow(log_data)
nrow_net_data <- nrow(net_data)
log_data <- log_data |> drop_na()
net_data <- net_data |> drop_na()
nrow_log_data - nrow(log_data)
nrow_net_data - nrow(net_data)
```

```{r}
nrow(log_data)
nrow(net_data)
```

## (c)

First check that location data is distinct on ID column:

```{r}
nrow(loc_data |> 
        select(ID) |>
        distinct()) == nrow(loc_data)
```

286700/292786 rows remain after joining for log data.

```{r}
log_data_loc <- merge(log_data, loc_data, by.x = "nodeid", by.y = "ID")
nrow(log_data)
nrow(log_data_loc)
```

110433/110718 rows remain after joining for net data.

```{r}
net_data_loc <- merge(net_data, loc_data, by.x = "nodeid", by.y = "ID")
nrow(net_data)
nrow(net_data_loc)
```

## (d)

Some of the histograms of the raw data together with Q1/median/Q3 without handling outliers:

```{r}
# par(mfcol = c(2, 5))
for (var in variables) {
  log_data_five_num <- fivenum(log_data_loc[[var]])
  names(log_data_five_num) <- c("min", "Q1", "median", "Q3", "max")
  plot1 <- ggplot(data = log_data_loc, aes_string(x = var)) +
           geom_histogram(na.rm = T, bins = 50) +
           geom_vline(xintercept = log_data_five_num["Q1"], linetype="dotted", color = "red") +
           geom_vline(xintercept = log_data_five_num["median"], linetype="dotted", color = "red") + 
           geom_vline(xintercept = log_data_five_num["Q3"], linetype="dotted", color = "red")
  
  net_data_five_num <- fivenum(net_data_loc[[var]])
  names(net_data_five_num) <- c("min", "Q1", "median", "Q3", "max")
  plot2 <- ggplot(data = net_data_loc, aes_string(x = var)) +
           geom_histogram(na.rm = T, bins = 50) +
           geom_vline(xintercept = net_data_five_num["Q1"], linetype="dotted", color = "red") +
           geom_vline(xintercept = net_data_five_num["median"], linetype="dotted", color = "red") + 
           geom_vline(xintercept = net_data_five_num["Q3"], linetype="dotted", color = "red")
  
  # print(plot1)
  # print(plot2)
  grid.arrange(plot1, plot2, nrow = 2)
  # plot3 <- grid.arrange(plot1, plot2, ncol = 2)
  # print(plot3)
}
```

The takeaways are that:
1. histograms are perfect tools for identifying existence of outliers.
2. No essential information can be drawn from histograms unless the outliers are handled properly.

One possible solution is to use 1.5 x IQR Rule.

```{r}
for (var in variables) {
  five_num <- fivenum(log_data_loc[[var]])
  Q1 <- five_num[2]
  Q3 <- five_num[4]
  IQR <- Q3 - Q1
  log_data_loc_clean <- log_data_loc |>
                          filter(!!as.symbol(var) >= Q1-1.5*IQR, !!as.symbol(var) <= Q3+1.5*IQR)
  five_num <- fivenum(log_data_loc_clean[[var]])
  Q1 <- five_num[2]
  median <- five_num[3]
  Q3 <- five_num[4]
  plot1 <- ggplot(data = log_data_loc_clean, aes_string(x = var)) +
           geom_histogram(na.rm = T, bins = 50) +
           geom_vline(xintercept = Q1, linetype = "dotted", color = "red") +
           geom_vline(xintercept = median, linetype = "dotted", color = "red") + 
           geom_vline(xintercept = Q3, linetype = "dotted", color = "red")
  
  five_num <- fivenum(net_data_loc[[var]])
  Q1 <- five_num[2]
  Q3 <- five_num[4]
  IQR <- Q3 - Q1
  net_data_loc_clean <- net_data_loc |>
                          filter(!!as.symbol(var) >= Q1-1.5*IQR, !!as.symbol(var) <= Q3+1.5*IQR)
  five_num <- fivenum(net_data_loc_clean[[var]])
  Q1 <- five_num[2]
  median <- five_num[3]
  Q3 <- five_num[4]
  plot2 <- ggplot(data = net_data_loc_clean, aes_string(x = var)) +
           geom_histogram(na.rm = T, bins = 50) +
           geom_vline(xintercept = Q1, linetype = "dotted", color = "red") +
           geom_vline(xintercept = median, linetype = "dotted", color = "red") + 
           geom_vline(xintercept = Q3, linetype = "dotted", color = "red")
  
  grid.arrange(plot1, plot2, nrow = 2)
}
```

This turns out to fail, some evidence:
1. For humidity in log data, values smaller than 0 and larger than 100 are not fully removed.
2. For hamabot in both log and net data, only values of 0 are kept because both Q1 and Q3 are 0.
Therefore, quantile-based outlier rejection is not working.

Second possible solution is to use hard-coded values. But it is very hard to determine the threshold for hamatop and hamabot: Values tend to spread out equally and evenly within reasonable ranges without sudden drop. The absence of sudden drop makes it hard to determine the proper threshold.

```{r}
# Variable voltage
plot1 <- log_data_loc |>
           filter(voltage >= 2) |>
           ggplot(aes(x = voltage)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data_loc |>
           filter(voltage >= 2) |>
           ggplot(aes(x = voltage)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

```{r}
# Variable humidity
plot1 <- log_data_loc |>
           filter(humidity >= 0, humidity <= 100) |>
           ggplot(aes(x = humidity)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data_loc |> 
           filter(humidity >= 0, humidity <= 100) |>
           ggplot(aes(x = humidity)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

```{r}
# Variable temperature
plot1 <- log_data_loc |>
           filter(humid_temp >= 0, humid_temp <= 40) |>
           ggplot(aes(x = humid_temp)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data_loc |> 
           filter(humid_temp >= 0, humid_temp <= 40) |>
           ggplot(aes(x = humid_temp)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

```{r}
# Variable hamatop
plot1 <- log_data_loc |>
           filter(hamatop <= 15000) |>
           ggplot(aes(x = hamatop)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data_loc |> 
           filter(hamatop <= 15000) |>
           ggplot(aes(x = hamatop)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

```{r}
# Variable hamabot
plot1 <- log_data_loc |>
           filter(hamabot <= 5000) |>
           ggplot(aes(x = hamabot)) +
           geom_histogram(na.rm = T, bins = 50)

plot2 <- net_data_loc |> 
           filter(hamabot <= 5000) |>
           ggplot(aes(x = hamabot)) +
           geom_histogram(na.rm = T, bins = 50)

grid.arrange(plot1, plot2, ncol = 2)
```

Examine the battery voltage when there are known anomalies:

```{r}
log_data_loc |>
  filter(humidity < 0 | humidity > 100 | humid_temp < 0 | humid_temp > 40 | hamatop > 15000 | hamabot > 5000) |>
  select(voltage) |>
  ggplot(aes(x = voltage)) +
  geom_histogram(bins = 50)
```

```{r}
net_data_loc |>
  filter(humidity < 0 | humidity > 100 | humid_temp < 0 | humid_temp > 40 | hamatop > 15000 | hamabot > 5000) |>
  select(voltage) |>
  ggplot(aes(x = voltage)) +
  geom_histogram(bins = 50)
```

It looks extremely abnormal. To confirm the guess, draw the histogram when readings are reasonable:

```{r}
log_data_loc |>
  filter(humidity >= 0, humidity <= 100, humid_temp >= 0, humid_temp <= 40, hamatop <= 15000, hamabot <= 5000) |>
  select(voltage) |>
  ggplot(aes(x = voltage)) +
  geom_histogram(bins = 50)
```

```{r}
log_data_loc |>
  filter(humidity >= 0, humidity <= 100, humid_temp >= 0, humid_temp <= 40, hamatop <= 2000, hamabot <= 200) |>
  select(voltage) |>
  ggplot(aes(x = voltage)) +
  geom_histogram(bins = 50)
```

Voltage still looks abnormal..........try something else.

```{r}
nrow(loc_data)
nrow(log_data_loc |> 
       select(nodeid) |>
       distinct())
nrow(net_data_loc |> 
       select(nodeid) |>
       distinct())
```

Almost all abnormal readings are generated by node 29 (this node is not functioning well throughout the experiment) except one generated by node 198.

```{r}
log_data_loc |>
  mutate(abnormal = case_when(humid_temp <= 0 ~ 1,
                              TRUE ~ 0)) |>
  dplyr::count(nodeid, abnormal)
```

```{r}
log_data_loc |>
  filter(nodeid == 29 | nodeid == 198) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))
```

```{r}
log_data_loc |>
  filter(nodeid == 198, humid_temp <= 0) |>
  select(voltage)
```

```{r}
log_data_loc |>
  filter(nodeid == 29 | nodeid == 198) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humid_temp)) +
  geom_line(aes(colour = nodeid))
```

```{r}
log_data_loc |>
  filter(voltage > 2.4) |>
  arrange(humid_temp)
```

```{r}
log_data_loc |>
  filter(voltage > 2.4) |>
  ggplot(aes(x = humid_temp)) +
  geom_histogram(bins = 50)
```

conclusion: abnormal readings are correlated to low voltage, 2.4 is a good threshold. After removing low-voltage readings, temperature readings fall inside normal range.

```{r}
net_data_loc |>
  mutate(abnormal = case_when(humid_temp >= 40 ~ 1,
                              TRUE ~ 0)) |>
  dplyr::count(nodeid, abnormal)
```

Selected nodeid from net_data: 3/78/123/141/145

```{r}
net_data_loc |>
  filter(nodeid == 78 | nodeid == 123 | nodeid == 145 | nodeid == 3 | nodeid == 141) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = result_time, y = humid_temp)) +
  geom_line(aes(colour = nodeid))
```

```{r}
net_data_loc |>
  filter(nodeid == 78 | nodeid == 123 | nodeid == 145 | nodeid == 3 | nodeid == 141) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = result_time, y = voltage)) +
  geom_line(aes(colour = nodeid))
```

Node 123/3/78 demonstrate the correlation mentioned in the thesis. Node 141 and 145 have problematic reading for voltage throughout the experiment. It's valid to drop readings with very low voltage. The next thing to validate is that if voltage reading is normal, temperature outliers are removed. It turns out that the dropping voltage readings larger than 3 is redundant. 2.4 itself is already a great threshold.

```{r}
net_data_loc |>
  filter(voltage > 2.4) |>
  arrange(desc(humid_temp))
```

Manually removing two more outliers gives clean distribution. Notice that two outliers come from problematic node 123.

```{r}
net_data_loc |>
  filter(voltage > 2.4, humid_temp < 100) |>
  ggplot(aes(x = humid_temp)) +
  geom_histogram(bins = 50)
```

Next is to check other readings when voltage is low.

It's interesting to see that low voltage correlates with negative humidity.

```{r}
log_data_loc |>
  filter(voltage <= 2.4) |>
  arrange(humidity)
```

If drop values by low voltage, quite a number of normal observations need to be dropped.

```{r}
log_data_loc |>
  filter(voltage <= 2.4, nodeid != 29, nodeid != 198) |>
  arrange(humidity)
```

```{r}
log_data_loc |>
  filter(voltage <= 2.4) |>
  distinct(nodeid)
```

```{r}
nodeid_log_low_voltage <- log_data_loc |>
  filter(voltage <= 2.4) |>
  group_by(nodeid) |>
  dplyr::summarise(avg_voltage = mean(voltage)) |>
  top_n(7, wt = -avg_voltage)
nodeid_log_low_voltage
```

Rest are normal.

```{r}
log_data_loc |>
  filter(nodeid != 29, nodeid != 198) |>
  arrange(humidity)
```

Rest of the readings are all normal for node 198.

```{r}
log_data_loc |>
  filter(nodeid == 198) |>
  arrange(humidity)
```

```{r}
net_data_loc |>
  filter(voltage <= 2.4) |>
  arrange(humidity)
```

```{r}
net_data_loc |>
  filter(voltage <= 2.4) |>
  arrange(desc(humidity))
```

```{r}
plot1 <- merge(log_data_loc, nodeid_log_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humidity)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(log_data_loc, nodeid_log_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

It turns out that dropping by low voltage is feasible for net data.

```{r}
net_data_loc |>
  filter(voltage <= 2.4, nodeid != 78, nodeid != 141, nodeid != 123, nodeid != 145, nodeid != 3) |>
  arrange(humidity)
```
Node 134 is actually OK in the later analysis but it is very weird to have voltage 0 reading. And there are very few observations for node 59 and 138. So it is feasible to drop by voltage.

```{r}
net_data_loc |>
  filter(voltage <= 2.4, nodeid != 78, nodeid != 141, nodeid != 123, nodeid != 145, nodeid != 3) |>
  arrange(desc(humidity))
```

Node 119 is problematic. 

```{r}
net_data_loc |>
  filter(nodeid != 78, nodeid != 141, nodeid != 123, nodeid != 145, nodeid != 3) |>
  arrange(humidity)
```

Node 118, 42, 105, 127 are problematic.

```{r}
net_data_loc |>
  filter(nodeid != 78, nodeid != 141, nodeid != 123, nodeid != 145, nodeid != 3) |>
  arrange(desc(humidity))
```

```{r}
nodeid_net_low_voltage <- net_data_loc |>
  filter(voltage <= 2.4) |>
  distinct(nodeid)
nodeid_net_low_voltage
```

```{r}
plot1 <- merge(net_data_loc, nodeid_net_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humidity)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc, nodeid_net_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

They are exhibiting similar trend so when the voltage is normal, the nodes are actually not problematic. We only need to remove outliers from node 118. Rest of the nodes turn out to have normal readings.

```{r}
plot1 <- net_data_loc |>
  filter(nodeid == 118 | nodeid == 42 | nodeid == 105 | nodeid == 127 | nodeid == 119) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humidity)) +
  geom_line(aes(colour = nodeid))

plot2 <- net_data_loc |>
  filter(nodeid == 118 | nodeid == 42 | nodeid == 105 | nodeid == 127 | nodeid == 119) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

The conclusion is that low voltage might be the cause of abnormal readings. When there are abnormal readings, the voltage is most of the time low. But the reverse is not true. When the voltage is low, the readings are not guaranteed to be abnormal. From the analysis of temperature/humidity and voltage, we see that it is more node-related than voltage-related. Moreover, for certain abnormal nodes, not all readings are abnormal. Abnormal readings tend to appear at the same time.

```{r}
log_data_loc |>
  filter(voltage <= 2.4) |>
  arrange(desc(hamatop))
```

Once again, only one reading is abnormal.

```{r}
log_data_loc |>
  filter(nodeid == 198) |>
  arrange(desc(hamatop))
```

For hamatop, the rest are all within normal ranges (low voltage).

```{r}
log_data_loc |>
  filter(voltage <= 2.4, nodeid != 198, nodeid != 29) |>
  arrange(desc(hamatop))
```

For hamatop, except for nodeid 40, everything else is within normal ranges.

```{r}
log_data_loc |>
  filter(nodeid != 198, nodeid != 29) |>
  arrange(desc(hamatop))
```

Investigate into node 40, readings are gone after only 100 epoch and abnormal readings are located in the later epochs. We'd better remove all of the reading from this node.

```{r}
log_data_loc |>
  filter(nodeid == 40) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(color = nodeid))
```

```{r}
nodeid_log_low_voltage <- log_data_loc |>
  filter(voltage <= 2.4) |>
  group_by(nodeid) |>
  dplyr::summarise(avg_voltage = mean(voltage)) |>
  arrange(avg_voltage) |>
  top_n(6, wt = -avg_voltage)
nodeid_log_low_voltage
```

Check readings for low voltage for hamatop. Everything works just fine.

```{r}
plot1 <- merge(log_data_loc, nodeid_log_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(log_data_loc, nodeid_log_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

When dealing with net data, remove all low voltage readings and outliers from node 118. All are within normal ranges.

```{r}
net_data_loc |>
  filter(voltage > 2.4, humidity < 110) |>
  arrange(desc(hamatop))
```

```{r}
nodeid_net_hamatop <- net_data_loc |>
  filter(voltage > 2.4, humidity < 110) |>
  group_by(nodeid) |>
  dplyr::summarise(avg_hamatop = mean(hamatop)) |>
  arrange(desc(avg_hamatop)) |>
  top_n(4, wt = avg_hamatop)
nodeid_net_hamatop
```

Net data now is absolutely clean on hamatop.

```{r}
merge(net_data_loc, nodeid_net_hamatop, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))
```

Finally, we deal with hamabot (for net). Everything looks fine. No outliers.

```{r}
net_data_loc |>
  filter(voltage > 2.4, humidity < 110) |>
  select(nodeid, hamabot) |>
  arrange(desc(hamabot))
```

```{r}
nodeid_net_hamabot <- net_data_loc |>
  filter(voltage > 2.4, humidity < 110) |>
  group_by(nodeid) |>
  dplyr::summarise(avg_hamabot = mean(hamabot)) |>
  arrange(desc(avg_hamabot)) |>
  top_n(4, wt = avg_hamabot)
nodeid_net_hamabot
```

Net data now is absolutely clean on hamabot.

```{r}
plot1 <- merge(net_data_loc, nodeid_net_hamabot, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamabot)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc, nodeid_net_hamabot, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2)
```

Next is hamabot for log data. First remove outliers by removing node 29/40 and one observation for node 198.

```{r}
log_data_loc |>
  filter(nodeid != 29, nodeid != 40, humidity >= -5000) |>
  select(nodeid, hamabot) |> 
  arrange(desc(hamabot))
```

Check readings for low voltage:

```{r}
plot1 <- merge(log_data_loc, nodeid_log_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamabot)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(log_data_loc, nodeid_log_low_voltage, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

Investigate into node 128. The trend is not reasonable. Remove node 128.

```{r}
plot1 <- log_data_loc |>
  filter(nodeid == 128) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(color = nodeid))

plot2 <- log_data_loc |>
  filter(nodeid == 128) |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(color = nodeid))

grid.arrange(plot1, plot2)
```

```{r}
nodeid_log_hamabot <- net_data_loc |>
  filter(nodeid != 29, nodeid != 40, humidity >= -5000) |>
  group_by(nodeid) |>
  dplyr::summarise(avg_hamabot = mean(hamabot)) |>
  arrange(desc(avg_hamabot)) |>
  top_n(2, wt = avg_hamabot)
nodeid_log_hamabot
```

Most of the time they are correlated. Looks fine.

```{r}
plot1 <- merge(log_data_loc, nodeid_log_hamabot, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamabot)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(log_data_loc, nodeid_log_hamabot, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2)
```

Check the summaries now (after merging with location data). We'll address each of the min/max for 4 variables.

```{r}
log_data_loc |>
  select(all_of(variables)) |>
  summary()
```

```{r}
net_data_loc |>
  select(all_of(variables)) |>
  summary()
```

All min values look fine now. We'll validate max values for humidity, hamatop and hamabot.

```{r}
log_data_loc_clean <- log_data_loc |>
  filter(nodeid != 29, nodeid != 40, nodeid != 128, humidity >= -5000)
log_data_loc_clean |>
  select(all_of(variables)) |>
  summary()
```

```{r}
net_data_loc_clean <- net_data_loc |>
  filter(voltage > 2.4, humidity < 110, humid_temp < 100)
net_data_loc_clean |>
  select(all_of(variables)) |>
  summary()
```

```{r}
log_data_loc_clean |>
  mutate(abnormal = case_when(humidity >= 100 ~ 1,
                              TRUE ~ 0)) |>
  dplyr::count(nodeid, abnormal) |>
  filter(abnormal == 1) |>
  select(nodeid, n) |>
  arrange(desc(n))
```

```{r}
nodeid_log_hum <- log_data_loc_clean |>
                    mutate(abnormal = case_when(humidity >= 100 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 1000) |>
                    select(nodeid) |>
                    distinct()
```

```{r}
plot1 <- merge(log_data_loc_clean, nodeid_log_hum, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humidity)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(log_data_loc_clean, nodeid_log_hum, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

```{r}
nodeid_log_hum <- log_data_loc_clean |>
                    mutate(abnormal = case_when(humidity >= 100 | humidity <= 0 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 900, n <= 1000) |>
                    select(nodeid) |>
                    distinct()
```

```{r}
plot1 <- merge(log_data_loc_clean, nodeid_log_hum, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humidity)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(log_data_loc_clean, nodeid_log_hum, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

```{r}
net_data_loc_clean |>
  mutate(abnormal = case_when(humidity >= 100 ~ 1,
                              TRUE ~ 0)) |>
  dplyr::count(nodeid, abnormal) |>
  filter(abnormal == 1) |>
  select(nodeid, n) |>
  arrange(desc(n))
```

```{r}
nodeid_net_hum <- net_data_loc_clean |>
                    mutate(abnormal = case_when(humidity >= 100 | humidity <= 0 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 1000) |>
                    select(nodeid) |>
                    distinct()

plot1 <- merge(net_data_loc_clean, nodeid_net_hum, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = humidity)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc_clean, nodeid_net_hum, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

Conclusion: there's no correlation between abnormal humidity reading and voltage reading for both datasets. Nodes tend to have similar trend. Therefore, those readings above 100 should be considered not accurate enough but they exhibit valid trend. Moreover, datasets have similar summary statistics. So we'll only consider net data for rest of the variables.

```{r}
net_data_loc_clean |>
  mutate(abnormal = case_when(hamatop >= 100000 ~ 1,
                              TRUE ~ 0)) |>
  group_by(nodeid, Height, abnormal) |>
  dplyr::summarise(n = n(), .groups = "drop") |>
  filter(abnormal == 1) |>
  select(nodeid, n, Height) |>
  arrange(desc(n))
```

```{r}
nodeid_net_top <- net_data_loc_clean |>
                    mutate(abnormal = case_when(hamatop >= 100000 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 400) |>
                    select(nodeid) |>
                    distinct()

plot1 <- merge(net_data_loc_clean, nodeid_net_top, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc_clean, nodeid_net_top, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

```{r}
nodeid_net_top <- net_data_loc_clean |>
                    mutate(abnormal = case_when(hamatop >= 100000 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 300, n <= 400) |>
                    select(nodeid) |>
                    distinct()

plot1 <- merge(net_data_loc_clean, nodeid_net_top, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc_clean, nodeid_net_top, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

```{r}
nodeid_net_top <- net_data_loc_clean |>
                    mutate(abnormal = case_when(hamatop >= 100000 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 150, n <= 300) |>
                    select(nodeid) |>
                    distinct()

plot1 <- merge(net_data_loc_clean, nodeid_net_top, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamatop)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc_clean, nodeid_net_top, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

The conclusion is that those very high readings are valid. They indeed reflects the actual incident light intensity. What's interesting is to notice that the node 138 in the last graph matches with the voltage correlation. The readings are gone after observing very low voltage.

```{r}
net_data_loc_clean |>
  mutate(abnormal = case_when(hamabot >= 5000 ~ 1,
                              TRUE ~ 0)) |>
  group_by(nodeid, Height, abnormal) |>
  dplyr::summarise(n = n(), .groups = "drop") |>
  filter(abnormal == 1) |>
  select(nodeid, n, Height) |>
  arrange(desc(n))
```

```{r}
nodeid_net_bottom <- net_data_loc_clean |>
                    mutate(abnormal = case_when(hamabot >= 5000 ~ 1,
                                                TRUE ~ 0)) |>
                    dplyr::count(nodeid, abnormal) |>
                    filter(abnormal == 1, n >= 30) |>
                    select(nodeid) |>
                    distinct()

plot1 <- merge(net_data_loc_clean, nodeid_net_bottom, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = hamabot)) +
  geom_line(aes(colour = nodeid))

plot2 <- merge(net_data_loc_clean, nodeid_net_bottom, by = "nodeid") |>
  mutate(nodeid = as.character(nodeid)) |>
  ggplot(aes(x = epoch, y = voltage)) +
  geom_line(aes(colour = nodeid))

grid.arrange(plot1, plot2, nrow = 2)
```

All 3 nodes exhibit same trend: they tend to reach peak at the same time. The conclusion is that the bottom readings are valid as well. 

Right now all min/max in the summary statistics for 4 variables are properly solved/explained. Final step is to check statistics for data cleaning. 10580/110433 removed for net data and 1597/286700 removed for log data.

```{r}
nrow(net_data_loc)
nrow(net_data_loc_clean)
nrow(net_data_loc) - nrow(net_data_loc_clean)
```

```{r}
nrow(log_data_loc)
nrow(log_data_loc_clean)
nrow(log_data_loc) - nrow(log_data_loc_clean)
```

Replace variables.

```{r}
net_data_loc <- net_data_loc_clean
log_data_loc <- log_data_loc_clean
```

## Part 3: Data Exploration

(a)

```{r}
net_data_loc |>
  dplyr::count(epoch) |>
  arrange(desc(epoch))
```

```{r}
log_data_loc |>
  dplyr::count(epoch) |>
  arrange(desc(epoch))
```


```{r}
outer <- merge(x = net_data_loc, y = log_data_loc, by = c("nodeid", "epoch"), all = T)
data_all <- outer |> 
                mutate( humidity = case_when(!is.na(humidity.x) ~ humidity.x,
                                    TRUE ~ humidity.y),
                        humid_temp = case_when(!is.na(humid_temp.x) ~ humid_temp.x,
                                            TRUE ~ humid_temp.y),
                        hamatop = case_when(!is.na(hamatop.x) ~ hamatop.x,
                                            TRUE ~ hamatop.y),
                        hamabot = case_when(!is.na(hamabot.x) ~ hamabot.x,
                                            TRUE ~ hamabot.y),
                        voltage = case_when(!is.na(voltage.x) ~ voltage.x,
                                            TRUE ~ voltage.y),
                        result_time = case_when(!is.na(result_time.x) ~ result_time.x,
                                            TRUE ~ result_time.y),
                        depth = case_when(!is.na(depth.x) ~ depth.x,
                                            TRUE ~ depth.y),
                        parent = case_when(!is.na(parent.x) ~ parent.x,
                                            TRUE ~ parent.y),
                        height = case_when(!is.na(Height.x) ~ Height.x,
                                            TRUE ~ Height.y),
                        direc = case_when(!is.na(Direc.x) ~ Direc.x,
                                            TRUE ~ Direc.y),
                        dist = case_when(!is.na(Dist.x) ~ Dist.x,
                                            TRUE ~ Dist.y),
                        tree = case_when(!is.na(Tree.x) ~ Tree.x,
                                            TRUE ~ Tree.y)) |>
                select(-ends_with('.x'), -ends_with('.y'))
```

```{r}
nrow(outer)
nrow(data_all)
```

```{r}
data_all |>
  mutate(day = as.POSIXct(round(result_time, "days"))) |>
  dplyr::count(day) |>
  arrange(day) |>
  ggplot(aes(x = day, y = n)) +
  geom_point() +
  geom_vline(xintercept = as.numeric(as.POSIXct(c("2004-05-26")))) +
  labs(x = "Day", y = "Count of data points per day") +
  annotate("text", x = as.POSIXct(c("2004-05-31")), y = 10000,
           label = "May 26", size = 4) +
  theme_bw()
```

```{r}
data <- data_all |>
  filter(result_time < as.POSIXct("2004-05-27"))
nrow(data)
```

```{r}
fig1 <- data %>%
  filter(hamabot > 0, hamatop > 0) %>%
  ggplot(aes(x = hamatop, y = hamabot)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Incident PAR", y = "Reflected PAR") +
  theme_bw()

fig2 <- data %>%
  ggplot(aes(x = humid_temp, y = humidity)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Temperature", y = "Relative Humidity") +
  theme_bw()

grid.arrange(fig1, fig2, ncol = 2)
```

```{r}
fig3 <- data %>%
  ggplot(aes(x = hamatop, y = height)) +
  geom_point(na.rm = T) +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Incident PAR", y = "Node Height") +
  theme_bw()

fig4 <- data %>%
  ggplot(aes(x = hamatop, y = humid_temp)) +
  geom_point(na.rm = T) +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Incident PAR", y = "Temperature") +
  theme_bw()

fig5 <- data %>%
  ggplot(aes(x = hamatop, y = hamabot)) +
  geom_point(na.rm = T) +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Incident PAR", y = "Reflected PAR") +
  theme_bw()

fig6 <- data %>%
  ggplot(aes(x = hamatop, y = humidity)) +
  geom_point(na.rm = T) +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Incident PAR", y = "Humidity") +
  theme_bw()

grid.arrange(fig3, fig4, fig5, fig6, nrow = 2, ncol = 2)
```

(c)

```{r fig.width=15, fig.height=8}
library(lubridate)

temp_plot <- data %>%
  group_by(date = date(result_time), height) %>%
  dplyr::summarise(across(humid_temp, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = humid_temp, color = height, group = height)) +
  geom_line() +
  scale_x_date(date_breaks = "3 days", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Temperature", x = "", y = "", color = "Node Height") +
  theme_bw()

humidity_plot <- data %>%
  group_by(date = date(result_time), height) %>%
  dplyr::summarise(across(humidity, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = humidity, color = height, group = height)) +
  geom_line() +
  scale_x_date(date_breaks = "3 days", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Relative Humidity", x = "", y = "", color = "Node Height") +
  theme_bw()

incident_plot <- data %>%
  group_by(date = date(result_time), height) %>%
  dplyr::summarise(across(hamatop, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = hamatop, color = height, group = height)) +
  geom_line() +
  scale_x_date(date_breaks = "3 days", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Incident PAR", x = "", y = "", color = "Node Height") +
  theme_bw()

reflected_plot <- data %>%
  group_by(date = date(result_time), height) %>%
  dplyr::summarise(across(hamabot, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = hamabot, color = height, group = height)) +
  geom_line() +
  scale_x_date(date_breaks = "3 days", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Reflected PAR", x = "", y = "", color = "Node Height") +
  theme_bw()

grid.arrange(temp_plot, humidity_plot, incident_plot, reflected_plot, ncol = 2, nrow = 2)
```

```{r}
temp_plot <- data %>%
  group_by(date = hour(result_time), height) %>%
  dplyr::summarise(across(humid_temp, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = humid_temp, color = height, group = height)) +
  geom_line() +
  # scale_x_date(date_breaks = "2 hours", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Temperature", x = "", y = "", color = "Node Height") +
  theme_bw()

humidity_plot <- data %>%
  group_by(date = hour(result_time), height) %>%
  dplyr::summarise(across(humidity, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = humidity, color = height, group = height)) +
  geom_line() +
  # scale_x_date(date_breaks = "2 hours", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Relative Humidity", x = "", y = "", color = "Node Height") +
  theme_bw()

incident_plot <- data %>%
  group_by(date = hour(result_time), height) %>%
  dplyr::summarise(across(hamatop, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = hamatop, color = height, group = height)) +
  geom_line() +
  # scale_x_date(date_breaks = "2 hours", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Incident PAR", x = "", y = "", color = "Node Height") +
  theme_bw()

reflected_plot <- data %>%
  group_by(date = hour(result_time), height) %>%
  dplyr::summarise(across(hamabot, mean), .groups = "drop") %>%
  ggplot(aes(x = date, y = hamabot, color = height, group = height)) +
  geom_line() +
  # scale_x_date(date_breaks = "2 hours", limits = c(as_date("2004-04-28"), as_date("2004-05-27")), expand = c(0,0), date_labels = "%B %d") +
  scale_color_gradient(low = "orange", high = "blue") +
  labs(title = "Reflected PAR", x = "", y = "", color = "Node Height") +
  theme_bw()

grid.arrange(temp_plot, humidity_plot, incident_plot, reflected_plot, ncol = 2, nrow = 2)
```

(d) 

PCA

```{r}
pca <- data %>%
  select(humidity, humid_temp, hamatop, hamabot) %>% 
  prcomp(center = TRUE, scale = TRUE)
# screeplot
eigenvalues <- pca$sdev^2
eigs_cum <- cumsum(eigenvalues) /sum(eigenvalues)
ggplot() + 
  geom_point(aes(x = 1:length(eigenvalues), y = eigs_cum)) + 
  geom_line(aes(x = 1:length(eigenvalues), y = eigs_cum)) +
  labs(x = "first PCs", y = "fraction of total variance explained") +
  theme_bw()
```

## Part 4: Interesting findings

```{r}
# KMeans for humidity annd temp against height
height <- data %>%
  group_by(height) %>%
  summarise(across(c(humidity, humid_temp), mean))

set.seed(2022)
cluster <- height %>%
  select(-height) %>%
  kmeans(centers = 2, nstart = 10)

height1 <-height %>%
  mutate(cluster = factor(cluster$cluster)) %>%
  ggplot(aes(x = height, fill = cluster, group = cluster)) +
  geom_density(alpha = 0.7, color = FALSE) +
  labs(title = "", x = "Height (m)", y = "", fill = "Cluster") +
  scale_fill_viridis_d() +
  theme_bw()

# KMeans for hamatop and hamabot against height
height <- data %>%
  group_by(height) %>%
  summarise(across(c(hamatop, hamabot), mean))

set.seed(2022)
cluster <- height %>%
  select(-height) %>%
  kmeans(centers = 2, nstart = 10)

height2 <- height %>%
  mutate(cluster = factor(cluster$cluster)) %>%
  ggplot(aes(x = height, fill = cluster, group = cluster)) +
  geom_density(alpha = 0.7, color = FALSE) +
  labs(title = "", x = "Height (m)", y = "", fill = "Cluster") +
  scale_fill_viridis_d() +
  theme_bw()

grid.arrange(height1, height2, nrow = 1, ncol = 2)
```

```{r}
# KMeans for humidity and temp against hour
library(lubridate)
hour <- data %>%
  mutate(hour = hour(result_time)) %>%
  group_by(hour) %>%
  summarise(across(c(humidity, humid_temp), mean))

set.seed(2022)
cluster <- hour %>%
  select(-hour) %>%
  kmeans(centers = 2, nstart = 10)

hour1 <- hour %>%
  mutate(cluster = factor(cluster$cluster)) %>%
  ggplot(aes(x = hour, fill = cluster, group = cluster)) +
  geom_density(alpha = 0.7, color = FALSE) +
  labs(title = "", x = "Hour", y = "", fill = "Cluster") +
  scale_fill_viridis_d() +
  theme_bw()

# KMeans for hamatop and hamabot against hour
hour <- data %>%
  mutate(hour = hour(result_time)) %>%
  group_by(hour) %>%
  summarise(across(c(hamatop, hamabot), mean))

set.seed(2022)
cluster <- hour %>%
  select(-hour) %>%
  kmeans(centers = 2, nstart = 10)

hour2 <- hour %>%
  mutate(cluster = factor(cluster$cluster)) %>%
  ggplot(aes(x = hour, fill = cluster, group = cluster)) +
  geom_density(alpha = 0.7, color = FALSE) +
  labs(title = "", x = "Hour", y = "", fill = "Cluster") +
  scale_fill_viridis_d() +
  theme_bw()

grid.arrange(hour1, hour2, nrow = 1, ncol = 2)
```

```{r}
# Show that the trend for temperature versus humidity is similar across chocices we choose
data %>%
  mutate(hour = hour(result_time),
         date = date(result_time)) %>%
  filter(date == "2004-05-06" | date == "2004-05-07" | date == "2004-05-08") %>%
  mutate(date = as.factor(date)) %>%
  ggplot(aes(x = humid_temp, y = humidity, color = date)) +
  geom_point(na.rm = T) +
  # geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(title = "", x = "Temperature", y = "Humidity") +
  theme_bw()
```


```{r}
# GMM
library(mixtools)
df <- data[as_date(data$result_time) == "2004-05-07",]
em <- mvnormalmixEM(df[, c("humidity", "humid_temp")])
plot(em, whichplots = 2)
```

```{r}
df$gmm_group_1 <- em$posterior[, 1] > 0.5 # T for group 1, F for group 2
df %>%
  ggplot() +
  geom_point(aes(x = humidity, y = humid_temp, group = gmm_group_1, color = gmm_group_1))
df %>%
  ggplot() +
  geom_point(aes(x = height, y = result_time, group = gmm_group_1, color = gmm_group_1))
```

```{r}
library(mclust)
set.seed(2022)
df <- data[as_date(data$result_time) == "2004-05-01",]
df_scaled <- df %>%
  select(hamatop, hamabot) %>%
  mutate(hamatop = scale(hamatop),
         hamabot = scale(hamabot))
model_gmm <- Mclust(df_scaled, G = 2)
z_gmm <- apply(model_gmm$z, 1, which.max)
df$cluster <- as.factor(z_gmm)
```

```{r}
df %>%
  ggplot() +
  geom_point(aes(x = hamatop, y = hamabot, group = cluster, color = cluster))
df %>%
  ggplot() +
  geom_point(aes(x = height, y = result_time, group = cluster, color = cluster))
```

```{r}
# PCA
library(ggfortify)
autoplot(pca, col = "hamabot", label = FALSE, loadings.label = TRUE) +
  theme_classic()
```

```{r}
# HC
library(cluster)
data = cbind(data, data.matrix(pca$x)[, c(1, 2)])
df1 <- data[as_date(data$result_time) == "2004-05-07", c("PC1", "PC2")]
m <- c("average", "single", "complete", "ward")
names(m) <- c("average", "single", "complete", "ward")
agg_coeff <- function(x){
  agnes(df1, method = x)$ac
}
result <- sapply(m, agg_coeff)
# ward has the highest agglomerative coeff
```

```{r}
cluster <- df1 %>%
  agnes(method = "ward")
pltree(cluster, cex = 0.6, hang = -1, main = "Dendogram") # plot the tree
# apply gap statistic to compare total intra-cluster variation
gap_stat <- clusGap(df1, FUN = hcut, nstart = 25, K.max = 10, B = 50)
# choose k with the highest gap statistic
fviz_gap_stat(gap_stat)
```

```{r}
final_clust <- hclust(dist(df1, method = "euclidean"), method = "ward.D2")
agnes_group <- cutree(final_clust, k = 5)
df1 <- data.frame(cbind(net_data_loc[as_date(net_data_loc$result_time) == "2004-05-07", ], agnes_group = as.factor(agnes_group)))
df1 %>% 
  ggplot() +
  geom_point(aes(result_time, Height, group = agnes_group, color = agnes_group), alpha = 0.5)
df1[df1$agnes_group == 4|df1$agnes_group == 5, ] %>%
  ggplot() +
  geom_point(aes(humid_temp, humidity, group = agnes_group, color = agnes_group), alpha = 0.5)
```

```{r}
df1[df1$agnes_group == 4 & df1$humid_temp < 11 & df1$humidity > 95, ]$hamabot
df1[df1$agnes_group == 4 & df1$humid_temp < 11 & df1$humidity > 95, ]$hamatop
# doesn't have low temperature and humidity
# low_temp & high humidity in 4: hamabot all between 20-40; hamatop all 0
unique(df1[df1$agnes_group == 4, ]$Direc)
unique(df1[df1$agnes_group == 5, ]$Direc)
```

## Part 5 Graph Critique

(a)

```{r}
plot1 <- data %>% 
  filter(hamatop > 0) %>%
  mutate(hamatop = log(hamatop)) %>%
  ggplot() +
  geom_histogram(aes(x = hamatop), bins = 40) +
  labs(x = "mol/m2/s", y = "counts", title = "Incident PAR")

plot2 <- data %>% 
  filter(hamabot > 0) %>%
  mutate(hamabot = log(hamabot)) %>%
  ggplot() +
  geom_histogram(aes(x = hamabot), bins = 40) +
  labs(x = "mol/m2/2", y = "counts", title = "Reflected PAR")

grid.arrange(plot1, plot2, ncol = 2)
```

(b)

```{r, fig.height=4}
box1 <- data %>%
  mutate(height = as.factor(height)) %>%
  ggplot(aes(x = humid_temp, y = height)) +
  geom_boxplot(outlier.colour = "black", outlier.size = 0.5, notch = T) +
  labs(title = "Temperature", x = "°C") +
  theme_bw()

box2 <- data %>%
  mutate(height = as.factor(height)) %>%
  ggplot(aes(x = humidity, y = height)) +
  geom_boxplot(outlier.colour = "black", outlier.size = 0.5, notch = T) +
  labs(title = "Humidity", x = "%RH") +
  theme_bw()

box3 <- data %>%
  mutate(height = as.factor(height)) %>%
  ggplot(aes(x = hamatop, y = height)) +
  geom_boxplot(outlier.colour = "black", outlier.size = 0.5, notch = F) +
  labs(title = "Incident PAR", x = "μmol/m2/s") +
  theme_bw()

box4 <- data %>%
  mutate(height = as.factor(height)) %>%
  ggplot(aes(x = hamabot, y = height)) +
  geom_boxplot(outlier.colour = "black", outlier.size = 0.5, notch = F) +
  labs(title = "Reflected PAR", x = "μmol/m2/s") +
  theme_bw()

grid.arrange(box1, box2, box3, box4, ncol = 4)
```

```{r}
avg_temp <- data %>%
  group_by(epoch) %>%
  summarise(avg_temp = mean(humid_temp))

box1 <- data %>%
  merge(avg_temp, by = "epoch") %>%
  mutate(height = as.factor(height),
         diff = humid_temp - avg_temp) %>%
  ggplot(aes(x = diff, y = height)) +
  geom_boxplot(outlier.colour = "black", outlier.size = 0.5, notch = T) +
  labs(title = "Temperature") +
  theme_bw()

avg_hum <- data %>%
  group_by(epoch) %>%
  summarise(avg_hum = mean(humidity))

box2 <- data %>%
  merge(avg_hum, by = "epoch") %>%
  mutate(height = as.factor(height),
         diff = humidity - avg_hum) %>%
  ggplot(aes(x = diff, y = height)) +
  geom_boxplot(outlier.colour = "black", outlier.size = 0.5, notch = T) +
  labs(title = "Humidity") +
  theme_bw()

grid.arrange(box1, box2, ncol = 2)
```


```{r}
hour <- as.integer(format(data$result_time, "%H"))
data$Period <- ifelse(hour < 6, "0-6",
                      ifelse(hour < 13, "6-13", ifelse(hour < 20, "13-20", "20-24")))
data$Day <- format(data$result_time, "%b %d")
```

```{r, fig.height=4}
# humidity
humidity.period.mean <- data %>%
  select(Period, humidity, height) %>%
  group_by(Period, height) %>%
  mutate(mean = mean(humidity)) %>%
  select(Period, height, mean) %>%
  unique()
# mean by period
fig1 <- humidity.period.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Period, color = Period), size = 0.3) +
  labs(y = "humidity")
# mean by days
humidity.day.mean <- data %>%
  select(Day, humidity, height) %>%
  group_by(Day, height) %>%
  mutate(mean = mean(humidity)) %>%
  select(Day, height, mean) %>%
  unique() %>%
  arrange(Day)
fig2 <- humidity.day.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Day, color= Day), size = 0.3, 
            show.legend = FALSE) +
  labs(y = "humidity")

# temperature
temp.period.mean <- data %>%
  select(Period, humid_temp, height) %>%
  group_by(Period, height) %>%
  mutate(mean = mean(humid_temp)) %>%
  select(Period, height, mean) %>%
  unique()
# mean by period
fig3 <- temp.period.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Period, color = Period), size = 0.3) +
  labs(y = "temperature")
# mean by days
temp.day.mean <- data %>%
  select(Day, humid_temp, height) %>%
  group_by(Day, height) %>%
  mutate(mean = mean(humid_temp)) %>%
  select(Day, height, mean) %>%
  unique() %>%
  arrange(Day)
fig4 <- temp.day.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Day, color= Day), size = 0.3,
            show.legend = FALSE) +
  labs(y = "temperature")

# hamatop
hamatop.period.mean <- data %>%
  select(Period, hamatop, height) %>%
  group_by(Period, height) %>%
  mutate(mean = mean(hamatop)) %>%
  select(Period, height, mean) %>%
  unique()
# mean by period
fig5 <- hamatop.period.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Period, color = Period), size = 0.3) +
  labs(y = "Incident PAR")
# mean by days
hamatop.day.mean <- data %>%
  select(Day, hamatop, height) %>%
  group_by(Day, height) %>%
  mutate(mean = mean(hamatop)) %>%
  select(Day, height, mean) %>%
  unique() %>%
  arrange(Day)
fig6 <- hamatop.day.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Day, color= Day), size = 0.3,
            show.legend = FALSE) +
  labs(y = "Incident PAR")

# hamabot
hamabot.period.mean <- data %>%
  select(Period, hamabot, height) %>%
  group_by(Period, height) %>%
  mutate(mean = mean(hamabot)) %>%
  select(Period, height, mean) %>%
  unique()
# mean by period
fig7 <- hamabot.period.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Period, color = Period), size = 0.3) +
  labs(y = "Reflected PAR")
# mean by days
temp.day.mean <- data %>%
  select(Day, hamabot, height) %>%
  group_by(Day, height) %>%
  mutate(mean = mean(hamabot)) %>%
  select(Day, height, mean) %>%
  unique() %>%
  arrange(Day)
fig8 <- temp.day.mean %>%
  ggplot() +
  geom_line(mapping = aes(x = height, y = mean, group = Day, color= Day), size = 0.3,
            show.legend = FALSE) +
  labs(y = "Reflected PAR")
grid.arrange(fig1, fig2, fig3, fig4, fig5, fig6, fig7, fig8, nrow = 4, ncol = 2)
```

(c)

```{r, fig.width=6}
library(reshape2)
# humid
humid.height.mean <- data[data$Day == "May 01", ] %>%
  select(result_time, humidity, height) %>%
  mutate(height_group = cut(height, 5)) %>%
  group_by(result_time, height_group) %>%
  mutate(mean = mean(humidity)) %>%
  select(result_time, height_group, mean) %>%
  unique()

fig1 <- humid.height.mean %>%
  melt(id.vars = c("result_time", "height_group")) %>%
  ggplot() +
  geom_line(mapping = aes(x = result_time, y = value, group = height_group, color = height_group), size = 0.3) +
  scale_color_brewer(type = "equal", palette = "Set2") +
  labs(x = "Time", y = "Humidity", color = "height")
# temperature
temp.height.mean <- data[data$Day == "May 01", ] %>%
  select(result_time, humid_temp, height) %>%
  mutate(height_group = cut(height, 5)) %>%
  group_by(result_time, height_group) %>%
  mutate(mean = mean(humid_temp)) %>%
  select(result_time, height_group, mean) %>%
  unique()

fig2 <- temp.height.mean %>%
  melt(id.vars = c("result_time", "height_group")) %>%
  ggplot() +
  geom_line(mapping = aes(x = result_time, y = value, group = height_group, color = height_group), size = 0.3) +
  scale_color_brewer(type = "equal", palette = "Set2") +
  labs(x = "Time", y = "Temperature", color = "height")
# hamatop
hamatop.height.mean <- data[data$Day == "May 01", ] %>%
  select(result_time, hamatop, height) %>%
  mutate(height_group = cut(height, 5)) %>%
  group_by(result_time, height_group) %>%
  mutate(mean = mean(hamatop)) %>%
  select(result_time, height_group, mean) %>%
  unique()

fig3 <- hamatop.height.mean %>%
  melt(id.vars = c("result_time", "height_group")) %>%
  ggplot() +
  geom_line(mapping = aes(x = result_time, y = value, group = height_group, color = height_group), size = 0.3) +
  scale_color_brewer(type = "equal", palette = "Set2") +
  labs(x = "Time", y = "Incident PAR", color = "height")
# hamabot
hamabot.height.mean <- data[data$Day == "May 01", ] %>%
  select(result_time, hamabot, height) %>%
  mutate(height_group = cut(height, 5)) %>%
  group_by(result_time, height_group) %>%
  mutate(mean = mean(hamabot)) %>%
  select(result_time, height_group, mean) %>%
  unique()

fig4 <- hamabot.height.mean %>%
  melt(id.vars = c("result_time", "height_group")) %>%
  ggplot() +
  geom_line(mapping = aes(x = result_time, y = value, group = height_group, color = height_group), size = 0.3) +
  scale_color_brewer(type = "equal", palette = "Set2") +
  labs(x = "Time", y = "Reflected PAR", color = "height")
grid.arrange(fig1, fig2, fig3, fig4, ncol = 2, nrow = 2)
```

(d)

```{r}
log_data_loc$category = "log"
net_data_loc$category = "net"
all <- rbind(log_data_loc, net_data_loc)
```

```{r}
library(scales)
all %>%
  select(Height, result_time, category) %>%
  unique() %>%
  ggplot() +
  geom_point(aes(x = as.POSIXct(result_time), y = as.factor(Height), color = category, group = category), size = 0.01, position = position_dodge(width = 0.6)) +
  labs(x = "Days", y = "Height") +
  scale_x_datetime(breaks = date_breaks("5 day"))
```